---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


Introduction:

In this report, we analyze the NYPD Shooting Incident dataset to uncover insights regarding shooting
incidents in New York City. We will visualize trends over the years and analyze the demographics of
individuals involved in these incidents.


install.packages("installr")
install_tinytex()
library(installr)
updateR()
suppressWarnings({
install.packages("knitr")
library(knitr)
install.packages("tinytex")
tinytex::install_tinytex()
})

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)

```
)
```{r}
url_NYPD <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

NYPD <- read.csv(url_NYPD)

```

```{r}

NYPD_clean <- NYPD %>%
  select(c("OCCUR_DATE","OCCUR_TIME","BORO","PRECINCT",
           "STATISTICAL_MURDER_FLAG","VIC_AGE_GROUP","VIC_SEX","VIC_RACE")) %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE),
         OCCUR_TIME = hms(OCCUR_TIME),
         STATISTICAL_MURDER_FLAG = as.logical(STATISTICAL_MURDER_FLAG),
         Shootings = 1,
         Year = year(OCCUR_DATE))


NYPD_clean %>%
  ggplot(aes(x = BORO, fill = BORO)) +
  geom_bar() +
  labs(title = "NYPD Shootings Incidents by Borough",
       subtitle = "(2006 - 2021)",
       x = "NYC Boroughs",
       y = "Total Number of Shootings",
       caption = "(Figure - 1)")

NYPD_clean %>%
  ggplot(aes(x = Year)) +
  geom_bar(fill = "lightblue",show.legend = FALSE) +
  labs(title = "NYPD Shootings Incidents by Year",
       x = "Years (2006 - 2021)",
       y = "Total Number of Shootings",
       caption = "(Figure - 2)")
```

```{r}
NYPD_year <- NYPD_clean %>%
  group_by(Year) %>%
  summarize(Shootings = sum(Shootings))

NYPD_year %>%
  ggplot(aes(x = as.factor(Year), y = Shootings)) +
  geom_line() +
  geom_point(color = "red") +
  scale_x_discrete(labels = as.character(2006:2021)) +
  labs(
    title = "NYPD Shooting Incidents by Year",
    x = "Years (2006 - 2021)",
    y = "Total Number of Shootings",
    caption = "(Figure - 3)"
  )

```



```{r}
NYPD_boro <- NYPD_clean %>%
  group_by(BORO, OCCUR_DATE,Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop') %>%
  select(BORO,OCCUR_DATE,Shootings,STATISTICAL_MURDER_FLAG) %>%
  ungroup()

NYPD_boro_year <- NYPD_clean %>%
  mutate(Year = year(OCCUR_DATE)) %>%
  group_by(BORO, Year, Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG), 
            .groups = 'drop') %>%
  select(BORO, Year,Shootings,STATISTICAL_MURDER_FLAG) %>%
  ungroup()

NYPD_boro_total <- NYPD_boro_year %>%
  group_by(BORO) %>%
  summarize(Shootings = sum(Shootings))
(7402 + 10365)/ sum(NYPD_boro_total$Shootings)
736/ sum(NYPD_boro_total$Shootings)
NYPD_boro_year %>%
  ggplot(aes(x = Year, y = Shootings,color = BORO)) +
  geom_line() +
  geom_point() +
  labs(title = "NYPD Shootings by Borough by Year",
       subtitle = "(2006 - 2021)",
       x = "Year",
       y = "Number of Shootings",
       caption = "(Figure - 4)")

```


```{r}
NYPD_boro <- NYPD_clean %>%
  group_by(BORO, OCCUR_DATE, Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop') %>%
  select(BORO, OCCUR_DATE, Shootings, STATISTICAL_MURDER_FLAG) %>%
  ungroup()

NYPD_boro %>%
  ggplot(aes(x = OCCUR_DATE, y = Shootings)) +
  geom_line() +
  scale_x_date(date_labels = "%Y %b") +
  labs(title = "NYPD Shootings Per Day",
       subtitle = "(2006 - 2021)",
       x = "Date",
       y = "Shootings per day",
       caption = "(Figure - 5)")

```



```{r}
NYPD_time_year <- NYPD_clean %>%
  mutate(Time_year = format(as.Date(OCCUR_DATE), "%m/%d")) %>%
  mutate(Time_year = as.Date(Time_year,"%m/%d")) %>%
  group_by(Time_year,Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop') %>%
  select(Time_year,Shootings,STATISTICAL_MURDER_FLAG) %>%
  ungroup()

NYPD_time_year %>% slice_max(Shootings, n = 2)
NYPD_July_5 <- NYPD_clean %>%
  mutate(Time_year = format(as.Date(OCCUR_DATE), "%m/%d"),
         Hour = hour(OCCUR_TIME)) %>%
  mutate(Time_year = as.Date(Time_year,"%m/%d")) %>%
  filter(Time_year == "2022-07-05") %>%
  group_by(Hour,Shootings) %>%
  summarize(Shootings = sum(Shootings),
            .groups = 'drop')


NYPD_time_year %>%
  ggplot(aes(x = Time_year, y = Shootings)) +
  geom_line() +
  geom_point(data = NYPD_time_year %>% slice_max(Shootings, n = 2),
             aes(color="Most Shootings")) +
  scale_x_date(date_labels = "%b") +
  labs(title = "NYPD Shootings on that Day",
       subtile = "(2006 - 2021)",
       colour = "",
       x = "Day",
       y = "Shootings on that day",
       caption = "(Figure - 6)")
```

### summary(clean)



```{r}
NYPD_time_day <- NYPD_clean %>%
  group_by(OCCUR_TIME,Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop') %>%
  select(OCCUR_TIME,Shootings,STATISTICAL_MURDER_FLAG)

NYPD_time_day %>%
  ggplot(aes(x = OCCUR_TIME, y = Shootings)) +
  geom_line() +
  scale_x_time() +
  labs(title = "NYPD Shootings by the Time of Day",
       subtitle = "(2006 - 2021)",
       x = "Time of day (24 Hour Time System)",
       y = "Shootings",
       caption = "(Figure - 7)")

NYPD_time_hour <- NYPD_clean %>%
  mutate(Hour = hour(OCCUR_TIME)) %>%
  group_by(Hour,Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop') %>%
  mutate(Hour2 = Hour^2) %>%
  select(Hour,Shootings,STATISTICAL_MURDER_FLAG, Hour2)

NYPD_time_hour_model <- lm(data = NYPD_time_hour, Shootings ~ Hour + Hour2)
summary(NYPD_time_hour_model)
NYPD_time_hour %>%
  ggplot(aes(x = Hour, y = Shootings)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x + I(x^2), linewidth = 1) +
  labs(title = "NYPD Shootings by Time of Day per Hour",
       subtitle = "(2006-2021)",
       x = " Hours (24 Hour Time System)",
       y = "Shootings that Hour",
       caption = "(Figure - 8)")
```


```{r}
# Plot shootings by gender
ggplot(NYPD_clean, aes(x = VIC_SEX)) +
  geom_bar(fill = "lightblue") +
  labs(title = "Shootings by Gender", x = "Gender", y = "Number of Shootings") +
  theme_minimal()
```
As we can see, there is a huge gender tilt vis-a-vis the gender of the victim.



MODEL
```{r}
###MODEL
# Check the unique values in the BOROUGH column
unique(NYPD_clean$BORO)


# Create a binary target variable for Brooklyn (1 = Yes, 0 = No)
NYPD_clean$BROOKLYN <- ifelse(NYPD_clean$BORO == "BROOKLYN", 1, 0)

# Build a logistic regression model
model <- glm(BROOKLYN ~ VIC_SEX + VIC_RACE, data = NYPD_clean, family = "binomial")

# Summarize the model
summary(model)

# Predict the probability of a shooting in Brooklyn
NYPD_clean$predicted_prob <- predict(model, type = "response")

# View the predicted probabilities
head(NYPD_clean$predicted_prob)


```


```{r}
NYPD_clean <- NYPD_clean %>%
 mutate(Population = case_when(
    BORO == "BROOKLYN" ~ 2576771,
    BORO == "QUEENS" ~ 2270976,
    BORO == "BRONX" ~ 1427056,
    BORO == "MANHATTAN" ~ 1629153,
    BORO == "STATEN ISLAND" ~ 475596,
    TRUE ~ NA_real_
  ))
head(NYPD_clean, 10)



nypd_shooting_rate <- NYPD_clean %>%
  group_by(BORO) %>%
  summarise(total_shootings = n(),
            population = unique(Population),
            shooting_rate = total_shootings / (population / 100000)) %>%
  arrange(desc(shooting_rate))

ggplot(nypd_shooting_rate, aes(x = BORO, y = shooting_rate)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  ggtitle("Chance of Getting Shot in Each Borough") +
  xlab("Borough") +
  ylab("Shooting Rate per 100,000 People") +
  theme_minimal()

```


```{r}
NYPD_clean <- NYPD_clean %>%
  mutate(Total = ifelse(!is.na(BORO), 1, 0)) %>%
  group_by(BORO) %>%
  mutate(Total = cumsum(Total))

lm_model <- lm(Total ~ Population, data = NYPD_clean)

summary(lm_model)


ggplot(NYPD_clean, aes(x = Population, y = Total)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Population") +
  ylab("Total Incidences") +
  ggtitle("Linear Regression: Total Incidences vs Population")

```

Biases and Conclusions :


There are a few types of biases we could potentially identify:

Geographic Bias: Some areas ( certain boroughs or neighborhoods) may have higher rates of shootings due to population density, police presence, or historical social factors. This could bias our model.

Temporal Bias: Certain periods ( holidays, specific years) might see more shootings due to various factors ( events), which could influence model predictions if not accounted for properly.

Gender Bias: The data may contain gender imbalances ( more incidents involving males), which could influence predictions of who is involved in shootings.


Conclusions :
The dataset offers crucial insights into the behavioral patterns, the victims, the race, locations etc, which may offer valuable insights for policy changes over time.

We can conclude that Bronx has the highest chance of getting shot per person compared to the other boroughs in New York City. Staten Island has the lowest chance of getting shot per person. However, it’s important to note that the difference in shooting rates between the boroughs is not very large, with the highest rate being only slightly above 0.5% and the lowest rate being just over 0.15%.

Based on the linear regression results, we can conclude that there is a positive relationship between the number of shooting incidents and the population size in each borough. In other words, as the population size increases, the number of shooting incidents tends to increase as well.


**We must always look for more insights because :
  Correlation does not imply causation and there may be other factors that contribute to the number of shooting incidents beyond just population size.!!